% ============================================================================
% Nathan Gopee — Résumé (Updated margins)
% Full LaTeX source — all info & comments preserved
% ============================================================================

% ----------------------------------------------------------------------------
%                     DOCUMENT CLASS & GLOBAL OPTIONS
% ----------------------------------------------------------------------------
\documentclass[letterpaper,11pt]{article}

% ----------------------------------------------------------------------------
%                                  PACKAGES
% ----------------------------------------------------------------------------
\usepackage{fontawesome5}  % icon fonts
\usepackage{latexsym}
\usepackage[empty]{fullpage}
\usepackage{titlesec}
\usepackage{marvosym}
\usepackage[usenames,dvipsnames]{color}
\usepackage{verbatim}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyhdr}
\usepackage[english]{babel}
\input{glyphtounicode}      

% ---------------------------  GEOMETRY  ------------------------------------
% Re-instated original small margins per user request
\usepackage[margin=0.5in, top=0.3in, bottom=0.5in]{geometry}
% ---------------------------------------------------------------------------
% Global spacing adjustments (kept identical)
% ---------------------------------------------------------------------------
\setlength{\parskip}{0pt}
\renewcommand{\baselinestretch}{0.95}\selectfont
\setlist[itemize,1]{itemsep=1pt, topsep=2pt, parsep=0pt, partopsep=0pt, leftmargin=0.15in}

% ------------------------------  HEADERS/FOOTERS  --------------------------
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}

% ------------------------------  SECTION STYLE  ----------------------------
\titleformat{\section}{\vspace{-8pt}\scshape\raggedright\large}{}{0em}{}[\color{black}\titlerule \vspace{-6pt}]

\pdfgentounicode=1

% ------------------------------  CUSTOM COMMANDS  --------------------------
\newcommand{\resumeItem}[1]{\item\scriptsize{#1}}
\newcommand{\resumeSubheading}[4]{
  \vspace{-2pt}\item
  \begin{tabular*}{0.97\textwidth}[t]{l@{\extracolsep{\fill}}r}
    \textbf{#1} & #2 \\
    \textit{#3} & \textit{#4} \\
  \end{tabular*}\vspace{-2pt}
}
% Project heading: title | icons, followed by small footnote line
\newcommand{\resumeProjectHeading}[4]{
  \vspace{-2pt}\item
  \begin{tabular*}{0.97\textwidth}{l@{\extracolsep{\fill}}r}
    \textbf{#1} \href{#4}{\faLink} & {\Large #2} \\
  \end{tabular*}\vspace{-2pt}
  {\small\faDatabase\ \hspace{-0.15in} {\tiny #3}}\\[-10pt]
}
\newcommand{\resumeSubHeadingListStart}{\begin{itemize}[label={},parsep=0pt]}
\newcommand{\resumeSubHeadingListEnd}{\end{itemize}}

% ============================================================================
%                                DOCUMENT BODY
% ============================================================================
\begin{document}

% ---------------------------------  HEADER  ---------------------------------
\begin{center}\vspace{-6pt}
  {\Huge\bfseries Nathan Gopee}\\
  {\scriptsize\faEnvelope\ ndg8743@gmail.com \quad|\quad\faPhone*\ (845) 853-5418 \quad|\quad\faMapMarker*\ New York \quad|\quad\faLinkedin\ linkedin.com/in/nathangopee \quad|\quad\faGithub\ github.com/ndg8743}\\[ -6pt]
\end{center}

% -------------------------------- EDUCATION ---------------------------------
\section{Education}
\resumeSubHeadingListStart
  \resumeSubheading{State University of New York at New Paltz}{New Paltz, NY}{M.S. Computer Science}{Jan. 2025 -- Present}
  \resumeSubheading{State University of New York at New Paltz}{New Paltz, NY}{B.S. Computer Science}{Aug. 2021 -- Dec. 2024}
\resumeSubHeadingListEnd

% -------------------------------- EXPERIENCE --------------------------------
\vspace{-8pt}
\section{Experience}
\resumeSubHeadingListStart
  \resumeSubheading{Graduate Teaching Assistant}{Jan. 2025 -- Present}{SUNY New Paltz}{New Paltz, NY}
    \begin{itemize}
      \resumeItem{Supported 50+ students across Assembly \& Computer Architecture and Principles of Programming Languages through weekly labs, office hours, and study sessions}
      \resumeItem{Graded assignments, quizzes, and projects; ran review sessions to strengthen comprehension and raise class averages by 10\%}
      \resumeItem{Integrated LCC.js (hydra.newpaltz.edu/lccjs) into lab sessions for interactive assembler/interpreter demos; collaborated with faculty to refine curriculum and develop study materials}
    \end{itemize}


  \resumeSubheading{Graduate Lab Manager}{Aug. 2024 -- Present}{SUNY New Paltz}{New Paltz, NY}
    \begin{itemize}
      \resumeItem{Maintained and refactored sites, implemented SSO auth/user management for \texttt{hydra.newpaltz.edu}}
      \resumeItem{Setup and maintain 50+ Linux systems, enforced cybersecurity policies, and creating internal software solutions}
      \resumeItem{Supported students and research infrastructure through debugging, tutoring, and lab maintenance}
      \resumeItem{Built and maintained several 3D printers, deployed a local LLM at \texttt{gpt.hydra.newpaltz.edu} on our ML servers}
      \resumeItem{Collaborated on network security protocols, redundancy/backups, and infrastructure expansion/redesign for the department}
    \end{itemize}
\resumeSubHeadingListEnd

% -------------------------------- PROJECTS ----------------------------------
\vspace{-8pt}
\section{Projects}
\resumeSubHeadingListStart

  \resumeProjectHeading{Git Learning Journey — Interactive Git Internals Platform}{React, React Flow, xterm.js, Node, Docker}{https://github.com/ndg8743/GG-git-learning}
    \begin{itemize}
      \resumeItem{Architected an interactive learning app visualizing Git’s underlying data structures (Trie, DAG, AVL, Skip‑List, Union‑Find, Bloom Filter).}
      \resumeItem{Built an in‑browser terminal and challenge‑validation workflow to teach real Git commands.}
      \resumeItem{Set up CI/CD with Docker to ensure reproducible environments and streamlined updates.}
    \end{itemize}

  \resumeProjectHeading{LCC.js — In‑browser Assembly Compiler/IDE}{React, Node, JavaScript/TypeScript, Docker}{https://github.com/ndg8743/lccjs}
    \begin{itemize}
      \resumeItem{Co‑developed a JS/TS re‑implementation of LCC (from Dos Reis’s textbook) as a Node package and web IDE.}
      \resumeItem{Created a React UI for live editing, assembling, context highlight/hover, and step‑through tutorials.}
      \resumeItem{Used in CPS 330 labs for teaching/working through LCC Assembly exercises.}
    \end{itemize}

  \resumeProjectHeading{QView3D — Advanced 3D‑Printer Farm Web App}{Python, Go, TypeScript, Vue, Flask, WebSockets, Docker}{https://github.com/sunyhydralab/QView3D}
    \begin{itemize}
      \resumeItem{Led full‑scale backend refactor and developed a Go‑based virtual printer emulator for safer G‑code testing.}
      \resumeItem{Extended compatibility beyond Prusa printers; refined USB serial communication and real‑time updates.}
      \resumeItem{Improved CI/CD and cross‑platform deploys (Windows/WSL, Linux, macOS).}
    \end{itemize}

  \resumeProjectHeading{Danmomo — Multiplayer Puzzle Game}{Godot 4.3, GDScript, Node, WebSockets, AWS}{https://github.com/ndg8743/Danmomo}
    \begin{itemize}
      \resumeItem{Implemented real‑time multiplayer fruit‑merging gameplay with synchronized server‑client state.}
      \resumeItem{Added bomb mechanics and leaderboard tracking; designed scalable backend on AWS.}
    \end{itemize}

  \resumeProjectHeading{Fitness Tracker — Full‑stack Web App}{Vue 3 (Composition API, TS), Node/Express, SQL, Bulma}{https://github.com/ndg8743/WSP-Final-Project-Fall2024}
    \begin{itemize}
      \resumeItem{Built exercise and meal logging with authentication, admin management, and RESTful endpoints.}
      \resumeItem{Implemented a responsive UI with Bulma and client‑side state via Composition API.}
    \end{itemize}

  \resumeProjectHeading{Personal Blog Portfolio — Web App}{Node/Express, SQL, HTML/CSS/JS}{https://github.com/ndg8743/PersonalSite}
    \begin{itemize}
      \resumeItem{Developed full‑stack site with user registration, project showcases, and blogging features.}
      \resumeItem{Implemented secure authentication and interactive UI components.}
    \end{itemize}

  \resumeProjectHeading{Fibonacci Fractal Engine}{Java, TypeScript, Vue, JUnit, Maven, Mockito, gRPC, Protobuf, Git}{https://github.com/ndg8743/Fibonacci-Fractal-Engine}
    \begin{itemize}
      \resumeItem{Led a 3‑person team to create a fractal generator with multi‑threaded computation.}
      \resumeItem{Implemented optimizations (HashMap caching, dynamic programming, gRPC) for >10\% speedup.}
    \end{itemize}
\resumeSubHeadingListEnd

  \resumeProjectHeading{Danmomo -- Multiplayer Puzzle Game}{\faGamepad\ \faNodeJs\ \faAws}{\faToolbox\ Godot (GDScript), Node, WebSockets, AWS, Git}{https://github.com/ndg8743/danmomo}
    \begin{itemize}
      \resumeItem{Developed a 2D multiplayer puzzle game with live AWS-hosted matchmaking with WebSockets}
      \resumeItem{Implemented real-time state sync and matchmaking on AWS Lambda}
    \end{itemize}

  \resumeProjectHeading{Git Learning Journey! -- Interactive Git Internals}{\faJs\ \faPython\ \faJava\ \faNodeJs\ \faDocker\ \faTerminal}{\faToolbox\ JavaScript, Python, Node.js, Docker, Git, Webpack, React Flow, Tailwind, SQLite}{https://hydra.newpaltz.edu/lccjs}
    \begin{itemize}
      \resumeItem{Designed React Flow learning graph with color-coded modules to teach students Git}
      \resumeItem{Implemented in-browser terminal emulator with xterm.js and challenge validation in Node}
      \resumeItem{Animated data-structure visualizations: Trie, DAG, AVL, Skip List, Union-Find, Bloom Filter}
    \end{itemize}

  \resumeProjectHeading{LCC.js -- In-browser Assembly IDE/Compiler}{\faJs\ \faDocker}{\faToolbox\ JavaScript, React, Tailwind, Webpack, Docker, HTML/CSS, SQLite}{https://hydra.newpaltz.edu/lccjs}
    \begin{itemize}
      \resumeItem{Collaborated on LCC.js and created an in-browser Assembly IDE/Compiler}
      \resumeItem{Built a full stack site with live code editing, assembling, context highlighting/hover and step-through tutorials}
      \resumeItem{Used in CPS 330 to teach assembly fundamentals and lab work}
    \end{itemize}

  \resumeProjectHeading{Homelab -- Fault-Tolerant Kubernetes Cluster}{\faNetworkWired\ \faDocker\ \faLinux}{\faToolbox\ Proxmox, k3s, Longhorn, OPNsense, Traefik, Wazuh, Prometheus, Grafana, Jellyfin}{N/A}
    \begin{itemize}
      \resumeItem In progress of a homelab, k3s on Proxmox + three GPU laptops, VIP fail-over, Longhorn 3-rep storage, NAS-backed NFS
      \resumeItem Automated media (*-arr*), SIEM (Wazuh) \textbf{and on-demand GPU jobs (LLM training/mining) via KEDA}, with zero-touch recovery if any node drops
    \end{itemize}

    %  \resumeProjectHeading{Fitness Tracker -- Web App}{\faGlobe\ \faVuejs\ \faJs\ \faNodeJs}{\faToolbox\ Vue, TypeScript, Node, Express, PostgreSQL, Supabase, RESTful, Bulma, Git}{https://github.com/ndg8743/fitness-tracker}
    %    \begin{itemize}
    %      \resumeItem{Built full-stack exercise, meal tracker and social media app}
    %    \end{itemize}

% ---------------------- CERTIFICATIONS & SKILLS -----------------------------
\vspace{-8pt}
\section{Certifications \& Skills}
\resumeSubHeadingListStart
  \resumeItem{\textbf{Certifications:} Systems Administration, C/C++ Development, Linux, Networking, Cloud Architecture, DevOps}
  \resumeItem{\textbf{Languages:} Java, C, C++, Python, Go, SQL, JavaScript/TypeScript, HTML/CSS, Assembly}
  \resumeItem{\textbf{Frameworks \& Tools:} Docker, Kubernetes, Git, CI/CD, Gradle, Flask, FastAPI, Vue, React, Express, Tailwind, Bulma, REST, WebSockets, HTTP, gRPC, WebHooks, Electron, PowerShell, SQL, AWS, Google Cloud, Oracle, Azure, Supabase, Pytorch, Jupyter, Pandas, NLTK}
\resumeSubHeadingListEnd

% ---------------------------------- END -------------------------------------
\end{document}
