\documentclass[11pt,letterpaper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\onehalfspacing

\usepackage{microtype}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{tabularx}
\usepackage{xcolor}
\usepackage{listings}
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    backgroundcolor=\color{gray!10},
    keywordstyle=\color{blue},
    commentstyle=\color{green!50!black},
    stringstyle=\color{red!70!black},
    numbers=left,
    numberstyle=\tiny\color{gray},
    numbersep=5pt,
}

\usepackage{amsmath}
\usepackage[hyphens]{url}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cleveref}
\usepackage[numbers,sort&compress]{natbib}

\title{
    \vspace{-1cm}
    \textbf{Update 12: USB4 RDMA Investigation} \\[0.5em]
    \large SoftRoCE over USB4 Networking for Distributed LLM Workloads
}

\author{
    Nathan Gopee \\
    \textit{MS Computer Science Candidate} \\
    SUNY New Paltz \\
    \texttt{gopeen1@newpaltz.edu}
}

\date{February 2026}

\begin{document}

\maketitle

\begin{abstract}
\noindent
This update investigates the feasibility of RDMA communication over USB4 networking between the Phase~1 cluster nodes (Cerberus and Chimera). USB4 offers a 20~Gbps point-to-point link that could supplement or replace the current 2.5~GbE RDMA path. We present hardware reconnaissance, driver analysis, and a gap assessment that identifies the requirements for USB4 RDMA deployment.
\end{abstract}

\section{Motivation}

Test~1 established SoftRoCE-based RDMA between Cerberus and Chimera over 2.5~GbE Realtek interfaces, achieving 0.91~Gbps via \texttt{ib\_write\_bw} and 111.86~MB/s via UCX \texttt{tag\_bw}. While sufficient for protocol validation, the 2.5~GbE bottleneck limits practical tensor transfer throughput to $\sim$38\% of even the Realtek's theoretical line rate.

USB4 offers a potential 20~Gbps point-to-point link---an 8$\times$ bandwidth improvement over the current 2.5~GbE path. If SoftRoCE can be layered over a USB4 network interface, the resulting RDMA channel could approach the throughput of the 10~GbE NICs (Intel X710, Aquantia AQC107) without requiring direct NIC-to-NIC cabling.

\section{USB4 Architecture Overview}

USB4 is based on the Thunderbolt 3 protocol and provides:

\begin{itemize}
    \item \textbf{Tunneling architecture}: USB4 encapsulates multiple protocols (USB 3.x, DisplayPort, PCIe) within a single physical link.
    \item \textbf{IP networking}: The \texttt{thunderbolt-net} Linux kernel module creates a virtual Ethernet interface over the USB4/Thunderbolt tunnel, presenting a standard \texttt{thunderboltN} or \texttt{enpXsYuZuW} network device.
    \item \textbf{High MTU}: The \texttt{thunderbolt-net} driver supports MTU up to 65,520 bytes, dramatically reducing per-packet overhead compared to standard 1500-byte Ethernet.
    \item \textbf{Point-to-point}: USB4 networking is inherently point-to-point with no switching fabric, eliminating congestion from other traffic.
\end{itemize}

The RDMA stack layering for USB4 is:
\begin{center}
\begin{tabular}{@{}cl@{}}
\toprule
\textbf{Layer} & \textbf{Component} \\
\midrule
5 & Application (UCX \texttt{perftest}, NCCL) \\
4 & UCX / libibverbs \\
3 & SoftRoCE (\texttt{rdma\_rxe}) \\
2 & \texttt{thunderbolt-net} (virtual Ethernet) \\
1 & USB4 tunnel manager \\
0 & USB4 physical link (20 Gbps) \\
\bottomrule
\end{tabular}
\end{center}

\section{Hardware Reconnaissance}

\subsection{Cerberus: USB4 Controller Present}

\begin{table}[htbp]
\centering
\caption{Cerberus USB4/Thunderbolt hardware}
\label{tab:cerberus-usb4}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Property} & \textbf{Value} \\
\midrule
Controller & ASMedia ASM4242 \\
PCI ID & \texttt{1b21:2426} (host), \texttt{1b21:2425} (downstream) \\
Kernel module & \texttt{thunderbolt} (loaded) \\
Sysfs domain & \texttt{domain0}, device \texttt{0-0} \\
USB4 device name & ASM4242 \\
Kernel & 6.17.0-14-generic \\
\bottomrule
\end{tabular}
\end{table}

Cerberus has a fully functional USB4 controller. The \texttt{thunderbolt} kernel module loaded automatically, and the ASM4242 controller registered as \texttt{domain0} in \texttt{/sys/bus/thunderbolt/devices/}. The controller supports USB4 tunneling including IP networking via \texttt{thunderbolt-net}.

\subsection{Chimera: USB 3.2 Only}

\begin{table}[htbp]
\centering
\caption{Chimera USB controller hardware}
\label{tab:chimera-usb4}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Property} & \textbf{Value} \\
\midrule
Controller & ASMedia ASM3242 \\
PCI ID & \texttt{1b21:3242} \\
Protocol & USB 3.2 Gen 2$\times$2 (20 Gbps) \\
Kernel driver & \texttt{xhci\_hcd} \\
Thunderbolt & \textbf{Not detected} \\
Motherboard & ASRock TRX40 Creator (TR 3960X) \\
TB3 header & Present on board (unpopulated) \\
\bottomrule
\end{tabular}
\end{table}

Chimera's ASM3242 is a USB 3.2 host controller---not USB4/Thunderbolt. The \texttt{thunderbolt} kernel module loaded without error but detected no devices in \texttt{/sys/bus/thunderbolt/devices/}. The ASRock TRX40 Creator motherboard has an onboard Thunderbolt 3 header (Intel Titan Ridge compatible), but this requires an add-in card (AIC) that is not currently installed.

\subsection{Protocol Mismatch Analysis}

USB4 networking via \texttt{thunderbolt-net} requires a Thunderbolt/USB4 tunnel between both endpoints. The tunnel negotiation fails when:

\begin{enumerate}
    \item The remote device does not have a USB4/Thunderbolt controller (current situation).
    \item The cable does not support USB4 signaling (passive USB-C cables may lack Thunderbolt support).
    \item The Thunderbolt security policy rejects the remote device (managed by the \texttt{bolt} daemon).
\end{enumerate}

In our case, condition (1) is the blocking issue: Cerberus has USB4 (ASM4242) but Chimera has only USB 3.2 (ASM3242). No Thunderbolt tunnel can be established, and therefore no \texttt{thunderbolt-net} interface will appear.

\section{Attempted Configuration}

Despite the hardware mismatch, the following steps were attempted to verify the diagnosis:

\begin{lstlisting}[language=bash,caption={USB4 networking setup attempt}]
# Cerberus: Thunderbolt modules (already loaded)
sudo modprobe thunderbolt
sudo modprobe thunderbolt-net

# Verify controller
ls /sys/bus/thunderbolt/devices/
# Output: 0-0  domain0  (local controller only)

# Chimera: Thunderbolt modules loaded
sudo modprobe thunderbolt
sudo modprobe thunderbolt-net
# Module loaded but: no devices in sysfs

# Check for USB gadget alternative
ls /sys/class/udc/
# Empty -- no USB Device Controller available
# Desktop xHCI controllers support host mode only

# Attempted g_ncm (USB NCM gadget)
sudo modprobe g_ncm
# Module loaded but cannot bind without UDC
\end{lstlisting}

\paragraph{USB gadget networking.}
An alternative approach---USB NCM/ECM gadget networking---was also investigated. This requires one machine to act as a USB device (gadget) via a USB Device Controller (UDC). Desktop machines with standard xHCI host controllers do not support gadget mode; OTG or device-capable USB controllers are required. The \texttt{dummy\_hcd} virtual UDC module was not available on kernel 6.17.0-14.

\section{Expected Performance (Projected)}

Based on USB4 20~Gbps specifications and SoftRoCE overhead characteristics:

\begin{table}[htbp]
\centering
\caption{Projected USB4 RDMA performance vs.\ current 2.5~GbE baseline}
\label{tab:usb4-projected}
\begin{tabular}{@{}lrrl@{}}
\toprule
\textbf{Metric} & \textbf{2.5 GbE (actual)} & \textbf{USB4 20G (projected)} & \textbf{Improvement} \\
\midrule
Raw link rate & 2.5 Gbps & 20 Gbps & 8$\times$ \\
\texttt{ib\_write\_bw} & 0.91 Gbps & 10--15 Gbps & 11--16$\times$ \\
UCX \texttt{tag\_bw} (1 MB) & 0.89 Gbps & 8--12 Gbps & 9--13$\times$ \\
Latency (1 byte) & $>$100 $\mu$s & 15--50 $\mu$s & 2--7$\times$ \\
Max MTU & 1500 B & 65,520 B & 44$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key assumptions.}
\begin{itemize}
    \item MTU 65,520 bytes reduces per-packet overhead by $\sim$44$\times$ compared to standard Ethernet, significantly benefiting large tensor transfers.
    \item SoftRoCE CPU overhead remains the primary bottleneck; both machines have high-core-count CPUs (48 logical cores on Cerberus, 48 on Chimera) to absorb this.
    \item The USB4 link is point-to-point with no competing traffic, unlike the shared 2.5~GbE LAN.
    \item GPUDirect RDMA is not possible over SoftRoCE regardless of link speed; all transfers must transit system memory.
\end{itemize}

\section{Resolution Path}

\subsection{Option A: Thunderbolt 3 AIC for Chimera (Recommended)}

The ASRock TRX40 Creator has a dedicated Thunderbolt 3 header. Installing a compatible Thunderbolt 3 Add-In Card (e.g., ASRock Thunderbolt 3 AIC R2.0 or GC-Titan Ridge) would:

\begin{enumerate}
    \item Add an Intel Titan Ridge (JHL7540) Thunderbolt 3 controller to Chimera.
    \item Enable USB4/Thunderbolt tunnel negotiation with Cerberus's ASM4242.
    \item Create a \texttt{thunderbolt-net} interface on both machines.
    \item Allow SoftRoCE to bind to the high-speed tunnel interface.
\end{enumerate}

\textbf{Estimated cost:} \$30--60 for a Thunderbolt 3 AIC.

\subsection{Option B: Direct 10 GbE Cabling}

Both machines have 10~GbE NICs (Intel X710 on Cerberus, Aquantia AQC107 on Chimera). A direct DAC/SFP+ cable between them would provide:

\begin{itemize}
    \item 10~Gbps full-duplex bandwidth.
    \item Native iWARP RDMA on the X710 side (Aquantia side still requires SoftRoCE).
    \item No additional hardware purchase required if a 10GbE DAC cable is available.
\end{itemize}

\subsection{Option C: USB 3.2 Gadget Networking}

If a USB OTG-capable adapter or bridge cable is available, USB 3.2 Gen 2$\times$2 networking could provide 20~Gbps. However:
\begin{itemize}
    \item Requires either a USB bridge cable (e.g., Plugable USB3-TRAN) or USB OTG support.
    \item Standard desktop USB controllers do not support gadget mode.
    \item Performance would be limited by the USB 3.2 protocol overhead.
\end{itemize}

\section{Conclusions}

\begin{enumerate}
    \item \textbf{USB4 RDMA is architecturally sound}: SoftRoCE over \texttt{thunderbolt-net} is a valid approach that requires no kernel patches---only standard modules (\texttt{thunderbolt}, \texttt{thunderbolt-net}, \texttt{rdma\_rxe}).

    \item \textbf{Hardware gap identified}: Chimera lacks a USB4/Thunderbolt controller. The ASM3242 USB 3.2 controller cannot establish Thunderbolt tunnels. The TRX40 Creator's onboard TB3 header provides a clear upgrade path via an inexpensive AIC.

    \item \textbf{Projected 8--16$\times$ improvement}: USB4 20~Gbps with MTU 65,520 should deliver 10--15~Gbps RDMA bandwidth, compared to 0.91~Gbps on the current 2.5~GbE path.

    \item \textbf{Next step}: Acquire a Thunderbolt 3 AIC for Chimera, install it on the TRX40 Creator's TB3 header, then repeat the SoftRoCE/UCX test suite over the \texttt{thunderbolt-net} interface.
\end{enumerate}

\section{Setup Script (For Future Deployment)}
\label{sec:usb4-setup-script}

The following script will configure USB4 RDMA once both endpoints have compatible controllers:

\begin{lstlisting}[language=bash,caption={USB4 SoftRoCE setup script}]
#!/bin/bash
# usb4-rdma-setup.sh -- Run on BOTH machines
set -e

HOSTNAME=$(hostname)
if [ "$HOSTNAME" = "cerberus" ]; then
    MY_IP="10.10.10.1"; PEER_IP="10.10.10.2"
elif [ "$HOSTNAME" = "chimera" ]; then
    MY_IP="10.10.10.2"; PEER_IP="10.10.10.1"
fi

# Load modules
sudo modprobe thunderbolt
sudo modprobe thunderbolt-net
sudo modprobe rdma_rxe

# Authorize Thunderbolt device
sudo apt install -y bolt
for dev in $(boltctl list -a | grep -oP '[a-f0-9-]{36}'); do
    boltctl enroll "$dev" --policy auto 2>/dev/null || true
done

# Find thunderbolt-net interface
sleep 3
TB_IFACE=""
for iface in $(ls /sys/class/net/); do
    driver=$(readlink -f /sys/class/net/$iface/device/driver \
             2>/dev/null || echo "")
    if echo "$driver" | grep -q "thunderbolt-net"; then
        TB_IFACE=$iface; break
    fi
done
[ -z "$TB_IFACE" ] && { echo "No TB interface"; exit 1; }

# Configure network with max MTU
sudo ip link set "$TB_IFACE" mtu 65520
sudo ip addr flush dev "$TB_IFACE"
sudo ip addr add "$MY_IP/24" dev "$TB_IFACE"
sudo ip link set "$TB_IFACE" up

# Add SoftRoCE
sudo rdma link delete rxe0 2>/dev/null || true
sudo rdma link add rxe0 type rxe netdev "$TB_IFACE"

echo "Ready. Test: ping $PEER_IP"
echo "RDMA:  ib_write_bw -d rxe0 --report_gbits $PEER_IP"
echo "UCX:   UCX_NET_DEVICES=$TB_IFACE ucx_perftest $PEER_IP"
\end{lstlisting}

\bibliography{update12}
\bibliographystyle{plainnat}

\end{document}
